Labor Market Income and Substitution Effects Explain Contemporary Natural Selection

### 2020-12-19 why-natural-selection.Rmd
Abdel tweaks

---
title: "Labor Market Income and Substitution Effects Explain Contemporary Natural Selection"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Amsterdam UMC. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects and that are strongly related to the socio-economic system."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\bfseries\sffamily\LARGE}
  - \titleformat*{\subsection}{\bfseries\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Times
---

@@

Living organisms evolve through natural selection, in which allele frequencies
change in the population through differential reproduction rates. Studying the
mechanisms behind natural selection can help us better understand how
individual differences in complex traits and disease risk arise. Geneticists
have long hypothesized that natural selection is taking place in modern human
populations [@galton1869hereditary;@darwin1871the]; recent work confirms this 
using genome-wide analysis
[@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
particular, genetic variants associated with higher educational attainment are
being selected against, although effect sizes appear small.

@@

We find selection effects on many polygenic scores. Effects are largely consistent
across generations. The strength of natural selection on a polygenic score is
associated with that score's correlation with education and earnings: scores
that predict lower education and earnings are being selected for. Also, across
the board, polygenic scores have stronger relationships with fertility among
specific subgroups. Selection effects are stronger in groups with lower
income and less education, among younger parents, people not living with a partner,
and people with many lifetime sexual partners. Outside these groups, effects are
weaker and often statistically insignificant. In some subgroups, the
direction of selection is even reversed: polygenic scores predicting
higher education and earnings are associated with *higher* fertility.

@@

Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in the
sample by 5-year birth intervals. Several scores show consistent increases or
declines over this 30-year period, of the order of 5% of a standard
deviation. These changes could reflect natural selection within the UK
population, but also emigration, or ascertainment bias within the sample. 
Respondents have higher income and are better educated than the UK population, 
and they may also differ on other unobserved characteristics [@10.1093/aje/kwx246].
Since richer and educated people also live longer, this bias could also
increase with age.

@@

To test directly for natural selection, we regress respondents' polygenic scores
(PGS) on their number of children ($y_i$):

@@

These results could be driven by age, if older respondents have lower income
and are less educated, and also more subject to selection on polygenic
scores. However, if we rerun the regressions, interacting the polygenic
score with income category and also with a quadratic in age, the
interaction with income remains significant at 0.05/`r n_regs_ic` for
`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
the PGS with age of leaving full time education and a quadratic in age,
the interaction with age leaving FTE remains significant at
0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.

Selection effects are also different between men and women (Appendix Figure
\@ref(fig:plot-sexes)). Differences are particularly large for educational
attainment, height and MDD. Several polygenic scores for mental illness and
personality traits are more selected for (or less against) among women,
including major depressive disorder (MDD), schizophrenia and neuroticism, while
extraversion is more selected for among men. Scores for waist circumference and
waist-hip ratio are less selected for among women, and scores for educational
attainment are more selected against, though the difference is only significant
for the "EA3" educational attainment PGS of @lee2018gene . One possible reason
for these sex differences is that polygenic scores may affect fertility via
success in marriage markets, and men and women may value different
characteristics in these.

@@

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb)a shows effect sizes estimated
separately for each tercile of AFLB. Several effects are strikingly different
across terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several polygenic
scores for body measurements are selected against only among older mothers. The
correlation between effect sizes for the youngest and oldest terciles is `r
cor_age_flb`. As before, recall that polygenic scores may have two effects:
changing the number of children within each category, and changing the number of
children by shifting respondents between categories.

@@

To investigate this further, we regress *number of children* on polygenic scores
*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)b). In `r opp_signed` out
of `r length(score_names)` cases, effects change sign when controls are added.
The correlation between effect sizes controlling for AFLB, and raw effect sizes,
is `r cor_raw_flb`. Thus, selection effects seem to come through two opposing
channels: an effect on AFLB, and an opposite-signed effect on number of children
controlling for AFLB.

@@

Since UK Biobank subjects are not representative of the wider population, our
results suggest that naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using population data. We try three
alternative weighting schemes: weighting by geography, age and
presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects 
among the entire sample, estimated with the three weighting schemes. 
Mean effect sizes across all polygenic scores are
increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables. Since the Biobank
sample seems to be ascertained in ways that shrink estimates of selection
effects, we suspect that our weighted estimates are still conservative.[^purifying]

@@

res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/AFLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(1))
  ) %>% 
  rename(p = p.value) %>% 
  ggplot(aes(
    x = estimate, 
    y = term, 
    shape = Weights, 
    fill = Weights, 
    colour = p < 0.05/{{n_regs}}
  )) +
  geom_point(size = 1.5, stroke = 0.5) +
  my_vline +
  pval_scale() +
  scale_fill_manual(values = c(
    "None"              = "grey15", 
    "Geographical"      = "dodgerblue4",
    "Age/Qualification" = "steelblue2",
    "Age/Qual/AFLB"      = "darkseagreen4"
  )) +
  scale_shape_manual(values = c(
    "None" = "square filled", 
    "Geographical" = "circle filled",
    "Age/Qualification" = "circle filled", 
    "Age/Qual/AFLB" = "circle filled"
  )) + 
  labs(x = "Effect size", y = "")

@@

Although the direction of selection effects does not change between the
generations, there are other differences. Compared to residualized polygenic
scores, selection effects on unresidualized scores are about ten percent higher
on average for *n siblings*, whereas effects for *number of children* barely
change (Appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could be
because earlier fertility is driven more by geographically clustered deprivation
[e.g. via an insurance motive, @rendall1998old], which may correlate with the
broad-scale genetic variation captured by principal components, such as ancestry.

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. This can potentially be explained by the
economic theory of fertility [@becker1960economic; @willis1973new;
@becker1976child]. According to this theory, increases in a person's wage affect their
fertility via two opposing channels. There is an *income effect* by which
children become more affordable, like any other good. There is also a
*substitution effect*: since childrearing has a cost in time, the opportunity
cost of childrearing increases if one's market wage is higher. The income effect
would lead higher earners to have more children. The substitution effect would
lead higher earners to have fewer children. If so, then genetic variants which
affect earnings potential in the labour market may cause opposing effects on
fertility. The income effect will cause natural selection in favour of
earnings-increasing variants. The substitution effect will do the reverse.

@@

Lastly, it is often assumed that the substitution effect will dominate for lone
parents, or those in unstable relationships, since they have less opportunity to
share childcare responsibilities, while the income effect will dominate for
couples who are able to reap the gains from specialization. Indeed, US fertility
decreases faster with education among single mothers than married mothers
[@baudin2015fertility]. This can explain why earnings-increasing genetic variants 
decrease fertility more among single parents than among couples. The same logic 
could explain our results for lifetime number of sexual partners, if this is 
associated with relationship stability.

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
behaviour also does not explain why selection effects should work in the
opposite direction among older parents, whereas the economic theory can explain
this via income effects, as described above. We test the alternative theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040). While risk attitude is always a
highly significant predictor of number of children, it has little impact on the
effect size or significance of the polygenic scores. The median ratio of effect sizes
between regressions with and without controls is `r median(ratio)`; all scores
which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
when controlling for risk attitude. Thus, although risk attitude does predict
number of children, it appears not to mediate selection effects. Overall, we
believe that the economic theory is the most likely explanation.

@@

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean PGS for educational attainment (EA3) of children by household income group. Blue is actual. Grey is hypothetical in the absence of selection effects.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3`     = value
      )

@@

Abdel Abdellaoui is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 849200011 from The Netherlands Organisation for Health Research and Development. This study was conducted using UK Biobank resources under application number 40310.

@@

Excluding childless people in the parents' generation could bias our estimates.
To learn about this, we compare effect sizes excluding and including childless
people in the *current* generation. The correlation
between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
different scores are broadly similar whether the childless are counted or not. 
Unfortunately, however, absolute effect sizes are smaller when the
childless are excluded, for `r n_smaller` out of 33 scores; the median
percentage change is `r percent(median_pct_diff)`. The fact that childless
people have such a strong effect on estimates makes it hard to compare absolute
effect sizes across generations. In particular, since the parents' generation has
a different distribution of numbers of children, childless people may have had 
more or less effect in that generation.

@@

As an alternative approach, , we run
regressions interacting polygenic scores with birth year, median split at 1950
("early born" versus "late born"). We use both *number of children* and
*number of siblings* as a dependent variable. We weight using
age/qualification cells, and further adjust for selection in the
parents' generation (see above).

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term     = pretty_names(term),
        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var  = fct_inorder(dep.var),
        reg.type = fct_recode(reg.type, 
                               Residualized   = "controlled", 
                               Unresidualized = "raw"
                             )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline + 
        facet_wrap(vars(dep.var), scales = "free_y") +
        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered() +
        theme(
          axis.text.x = element_text(size = 8)
        )

@@

res_age_birth_parents %>% 
    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
    mutate(
      term    = pretty_names(term),
      term    = fct_reorder(term, estimate, order_abs(2)),
      Control = ifelse(control == "fath_age_birth", 
                  "Father's age at birth", 
                  "Mother's age at birth"
                )
    ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(y = term, x = estimate, fill = Control, 
      colour = p < 0.05/{{n_regs}})) + 
      pval_geom_point(alpha = 0.8) +  
      my_vline +
      my_fill_scale(2) +
      pval_scale() +
      my_labs()

@@

Our results suggest that polygenic scores may directly correlate with age at
first live birth. Figure \@ref(fig:plot-age-flb-dv) plots estimated effect sizes
from bivariate regressions for respondents, and Figure
\@ref(fig:plot-age-birth-parents-dv) does the same for their parents. Effect
sizes are reasonably large. They are also very highly correlated across
generations. Effect sizes of polygenic scores on father's age at own birth, and
on own age at first live birth, have a correlation of `r cor_flb_fath`; for
mother's age and own age it is `r cor_flb_moth`.

@@

res_age_birth_parents_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1)),
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's AFLB", 
          "moth_age_birth" = "Mother's AFLB"
        )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(estimate, term, fill = `Dependent variable`, 
        colour = p < 0.05 / {{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs()

@@

Another way to examine the "earnings" theory of natural selection is to
compare selection effects of PGS with their genetic correlation with
educational attainment (EA3). Since EA3 strongly predicts earnings, if
earnings drives differences in fertility, we'd expect a correlation
between the two sets of results. Figure
\@ref(fig:plot-rgs-by-effect-size) shows this is so.

@@

rgs %>% 
      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
      ggplot(aes(rg, estimate)) + 
        geom_point() +
        my_vline +
        my_hline +
        labs(x = "rG with EA3", y = "Effect on number of children")

@@

Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean
polygenic scores between the highest and lowest income groups. Column "Actual"
uses respondents' scores weighted by Age/Qualification times number of
children. Column "Without selection" uses scores weighted by Age/Qualification
only, i.e. if all couples had the same number of children.

@@

ineq_comp %>% 
      select(
        PGS                 = score, 
        `Without selection` = diff_hypoth,
        `With selection`    = diff_children, 
        `% change`          = pct_change
      ) %>% 
      mutate(
        PGS = pretty_names(PGS)
      ) %>% 
      arrange(desc(`% change`)) %>% 
      as_hux() %>% 
      set_align(everywhere, -1, "right") %>% 
      set_number_format(everywhere, 2:3, 3) %>% 
      set_number_format(everywhere, 4, 1) %>% 
      set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
      theme_compact() %>% 
      set_bottom_border(final(1), everywhere)
```

>>>>

---
title: "Labor Market Income and Substitution Effects Explain Contemporary Natural Selection"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among lower-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects and that are strongly related to the socio-economic system."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
sansfont: "Gill Sans"
---

@@

Living organisms evolve through natural selection, in which allele frequencies
change in the population through differential reproduction rates. Studying the
mechanisms behind natural selection can help us better understand how individual
differences in complex traits and disease risk arise. Recent work confirms that
natural selection is taking place in modern human populations, using genome-wide
analysis
[@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
particular, genetic variants associated with higher educational attainment are
being selected against, although effect sizes appear small.

@@

We find selection effects on many polygenic scores. Effects are largely consistent
across generations. The strength of natural selection on a polygenic score is
associated with that score's correlation with education and earnings: scores
that predict lower education and earnings are being selected for. Also, across
the board, polygenic scores have stronger relationships with fertility among
specific subgroups. Selection effects are stronger in groups with lower
income and less education, among younger parents, people not living with a partner,
and people with more lifetime sexual partners. Outside these groups, effects are
weaker and often statistically insignificant. In some subgroups, the
direction of selection is even reversed: polygenic scores predicting
higher education and earnings are associated with *higher* fertility.

@@

We created polygenic scores for 33 traits in `r nrow(famhist)` individuals,
corrected for ancestry using 100 genetic principal components (see Materials and
Methods). Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in
the sample by 5-year birth intervals. Several scores show consistent increases
or declines over this 30-year period, of the order of 5% of a standard
deviation. These changes could reflect natural selection within the UK
population, but also emigration, or ascertainment bias within the sample.
Respondents have higher income and are better educated than the UK population,
and they may also differ on other unobserved characteristics
[@10.1093/aje/kwx246]. Since richer and educated people also live longer, this
bias could also increase with age.

@@

To test directly for natural selection, we regress respondents' number of
children ($y_i$) on their polygenic scores (PGS):

@@

These results could be explained by age, if older respondents have lower income
and are less educated, and also show more natural selection on polygenic
scores. However, if we rerun the regressions, interacting the polygenic
score with income category and also with a quadratic in age, the
interaction with income remains significant at 0.05/`r n_regs_ic` for
`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
the PGS with age of leaving full time education and a quadratic in age,
the interaction with age leaving full time education remains significant at
0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.

Selection effects are also different between men and women (Appendix Figure
\@ref(fig:plot-sexes)). Differences are particularly large for educational
attainment, height and MDD. Several polygenic scores for mental illness and
personality traits are more selected for (or less against) among women,
including major depressive disorder (MDD), schizophrenia and neuroticism, while
extraversion is more selected for among men. Scores for waist circumference and
waist-hip ratio are less selected for among women, and scores for educational
attainment are more selected against, though the difference is only significant
for the ("EA3") educational attainment score of @lee2018gene. One
possible reason for these sex differences is that polygenic scores may affect
fertility via success in marriage markets, and men and women may value different
characteristics in these.

@@

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb)a shows effect sizes estimated
separately for each tercile of AFLB. Several effects are strikingly different
across terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several polygenic
scores for body measurements are selected against only among older mothers. The
correlation between effect sizes for the youngest and oldest terciles is 
`r cor_age_flb`.

@@

To investigate this further, we regress *number of children* on polygenic scores
*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)b). In `r opp_signed` out
of `r length(score_names)` cases, effects change sign when controls are added.
The correlation between effect sizes controlling for AFLB, and raw effect sizes,
is `r cor_raw_flb`. Thus, selection effects seem to come through two opposing
channels: an effect on AFLB, and an opposite-signed effect on number of children
controlling for AFLB. Note that we do not claim that AFLB is exogenous to an
individual's polygenic scores -- indeed, our analyses show that it is not
(Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue that AFLB mediates
part of the relationship between polygenic scores and fertility, and that
once this is controlled for, the remaining part of the relationship has the
opposite sign.

AFLB mediates
the relationship between polygenic scores and fertility, and that after this
effect has been taken account of, there is a further, opposite-signed relationship.

@@

Since UK Biobank participants are not representative of the wider population,
our results suggest that naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using population
data. We try three alternative weighting schemes: weighting by geography, age
and presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes.
Mean effect sizes across all polygenic scores are
increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables. Since the UK Biobank
sample seems to be ascertained in ways that shrink estimates of selection
effects, we suspect that our weighted estimates are still conservative.[^purifying]

@@

res_wt_combined %>% 
  mutate(
    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                "Age/Qual/AFLB", "None"),
    term     = pretty_names(term),
    term     = fct_reorder(term, estimate, order_abs(1))
  ) %>% 
  rename(p = p.value) %>% 
  ggplot(aes(
    x = estimate, 
    y = term, 
    shape = Weights, 
    fill = Weights, 
    colour = p < 0.05/{{n_regs}}
  )) +
  geom_point(size = 1.5, stroke = 0.5) +
  my_vline +
  pval_scale() +
  scale_fill_manual(values = c(
    "None"              = "grey15", 
    "Geographical"      = "dodgerblue4",
    "Age/Qualification" = "steelblue2",
    "Age/Qual/AFLB"      = "darkseagreen4"
  )) +
  scale_shape_manual(values = c(
    "None" = "square filled", 
    "Geographical" = "circle filled",
    "Age/Qualification" = "circle filled", 
    "Age/Qual/AFLB" = "circle filled"
  )) + 
  labs(x = "Effect size", y = "") +
  my_theme()

@@

Although the direction of selection effects does not change between the
generations, there are other differences. Compared to our standard polygenic
scores (residualized on 100 principal components of genetic data) 
selection effects on unresidualized scores are about ten percent higher
on average for *number of siblings*, whereas effects for *number of children*
barely change (Appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could
be because earlier fertility is driven more by geographically clustered
deprivation [e.g. via an insurance motive, @rendall1998old], which may correlate
with the broad-scale genetic variation captured by principal components, such as
ancestry.

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. One possible explanation for this is
given by the economic theory of fertility 
[@becker1960economic; @willis1973new;@becker1976child]. According to this theory,
increases in a person's wage affect their fertility via two opposing channels.
There is an *income effect* by which children become more affordable, like any
other good. There is also a *substitution effect*: since childrearing has a cost
in time, the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. If so,
then genetic variants which affect earnings potential in the labour market may
cause opposing effects on fertility. The income effect will cause natural
selection in favour of earnings-increasing variants. The substitution effect
will do the reverse.

@@

Lastly, it is often assumed that the substitution effect will dominate for lone
parents, or those in unstable relationships, since they have less opportunity to
share childcare responsibilities, while the income effect will dominate for
couples who are able to reap the gains from specialization [@becker1981treatise]. Indeed, US fertility
decreases faster with education among single mothers than married mothers
[@baudin2015fertility]. This can explain why earnings-increasing genetic variants 
decrease fertility more among single parents than among couples. The same logic 
could explain our results for lifetime number of sexual partners, if this is 
associated with relationship stability.

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures, e.g. waist-hip ratio and BMI.
Externalizing behaviour also does not explain why selection effects should work
in the opposite direction among older parents, whereas the economic theory can
explain this via income effects, as described above. We test the alternative
theory directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040). While risk attitude is always
a highly significant predictor of number of children, it has little impact on
the effect size or significance of the polygenic scores. The median ratio of
effect sizes between regressions with and without controls is `r median(ratio)`;
all scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. Thus, although risk attitude does
predict number of children, it appears not to mediate selection effects.
Overall, we believe that the economic theory is the most likely explanation.

@@

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group. Blue is actual. Grey is hypothetical in the absence of selection effects.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3`     = value
      )

@@

AA is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 849200011 from The Netherlands Organisation for Health Research and Development. This study was conducted using UK Biobank resources under application numbers 40310 and 19127.

@@

Excluding childless people in the parents' generation could bias our estimates.
To learn about this, we compare effect sizes excluding and including childless
people in the *current* generation. The correlation
between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
different scores are broadly similar whether the childless are counted or not. 
However, absolute effect sizes are smaller when the
childless are excluded, for `r n_smaller` out of 33 scores; the median
percentage change is `r percent(median_pct_diff)`. The fact that childless
people have such a strong effect on estimates makes it hard to compare total
effect sizes across generations. In particular, since the parents' generation has
a different distribution of numbers of children, childless people may have had 
more or less effect in that generation.

@@

As an alternative approach, we run regressions interacting polygenic scores with
birth year, median split at 1950 ("early born" versus "late born"). We use both
*number of children* and *number of siblings* as a dependent variable. We weight
using age/qualification cells, and further adjust for selection in the parents'
generation (see above).

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term     = pretty_names(term),
        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var  = fct_inorder(dep.var),
        reg.type = fct_recode(reg.type, 
                               Residualized   = "controlled", 
                               Unresidualized = "raw"
                             )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline + 
        facet_wrap(vars(dep.var), scales = "free") +
        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered() +
        theme(
          axis.text.x = element_text(size = 8)
        )

@@

res_age_birth_parents %>% 
    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
    mutate(
      term    = pretty_names(term),
      term    = fct_reorder(term, estimate, order_abs(2)),
      Control = ifelse(control == "fath_age_birth", 
                  "Father's age at birth", 
                  "Mother's age at birth"
                )
    ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(y = term, x = estimate, fill = Control, 
      colour = p < 0.05/{{n_regs}})) + 
      pval_geom_point(alpha = 0.8) +  
      my_vline +
      my_fill_scale(2) +
      pval_scale() +
      my_labs() +
      my_theme()

@@

Our results suggest that polygenic scores may directly correlate with age at
first live birth. Figure \@ref(fig:plot-age-flb-dv) plots estimated effect sizes
from bivariate regressions for respondents. Figure
\@ref(fig:plot-age-birth-parents-dv) does the same for their parents, using only
eldest siblings.[^only-eldest] Effect sizes are reasonably large. They are also
highly correlated across generations. Effect sizes of polygenic scores on
father's age at own birth, and on own age at first live birth, have a
correlation of `r cor_flb_fath`; for mother's age and own age it is 
`r cor_flb_moth`.

[^only-eldest]: Parental AFLB can only be calculated for this group.

@@

res_age_birth_parents_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1)),
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's AFLB", 
          "moth_age_birth" = "Mother's AFLB"
        )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(estimate, term, fill = `Dependent variable`, 
        colour = p < 0.05 / {{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() +
        my_theme()

@@

Another way to examine the "earnings" theory of natural selection is to compare
selection effects of PGS with their genetic correlation with educational
attainment (EA3). Since EA3 strongly predicts earnings, if earnings drives
differences in fertility, we'd expect a correlation between the two sets of
results. Figure \@ref(fig:plot-rgs-by-effect-size) shows this is so. Genetic
correlations were calculated using LD score regression from GWAS summary
statistics.

@@

rgs %>% 
      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
      ggplot(aes(rg, estimate)) + 
        geom_point() +
        my_vline +
        my_hline +
        labs(x = "Genetic correlation with EA3", y = "Effect on number of children")

@@

Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean
polygenic scores between the highest and lowest income groups. Column "With
selection" uses respondents' scores weighted by Age/Qualification times number
of children. Column "Without selection" uses scores weighted by
Age/Qualification only, i.e. if all couples had the same number of children.

@@

ineq_comp %>% 
      select(
        PGS                 = score, 
        `Without selection` = diff_hypoth,
        `With selection`    = diff_children, 
        `% change`          = pct_change
      ) %>% 
      mutate(
        PGS = pretty_names(PGS)
      ) %>% 
      arrange(desc(`% change`)) %>% 
      as_hux() %>% 
      set_align(everywhere, -1, "right") %>% 
      set_number_format(everywhere, 2:3, 3) %>% 
      set_number_format(everywhere, 4, 1) %>% 
      set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
      set_width(0.6) %>% 
      theme_compact() %>% 
      set_bottom_border(final(1), everywhere)
```
